Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-08-28 docs/manuscript/paper.tex
generate a different set of alphas
complete manuscript

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[Kimpson]{Kimpson$^{1}$, Melatos$^{2}$, O'Leary, Evans$^{3}$, others, etc. etc.%
\thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
%\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
\\
% List of institutions
$^{1}$School of Physics, University of Melbourne, Parkville, VIC 3010, Australia \\
$^{2}$OzGrav, University of Melbourne, Parkville, VIC 3010, Australia \\
$^{3}$Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia }

@@

% Abstract of the paper
\begin{abstract}	
Continuous nHz gravitational waves from individual supermassive black hole binaries are detectable with Pulsar Timing Array observations. A novel search strategy is developed wherein intrinsic pulsar timing variations are tracked simultaneously with the modulation induced by a putative gravitational wave. A two-step procedure is applied within a state-space framework, where the model parameters are estimated with a Kalman filter, which provides a likelihood for nested sampling. This enables self-consistent inference of model parameters and calculation of the Bayesian evidence. It is shown via an astrophysically representative software injection campaign that the procedure can distinguish a gravitational wave from pure noise down to a strain of $h_0 \sim 4 \times 10^{15}$, given an inclination angle $\iota = 1.0$ rad. Full posterior distributions of model parameters are recovered. There is a bias in the estimates of some parameters, introduced by dropping the so-called `pulsar terms', which is analysed and discussed.
\end{abstract}

@@

\section{State-Space Formulation}\label{sec:model}
We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation. In this work we take the intrinsic state variable to be the pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local rest frame of the pulsar's centre of mass. A model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.  We take the measurement variable to be the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$.  The measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_measured}. The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array.
\subsection{Spin evolution} \label{sec:psr_frequency}

A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}. We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend. In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque. The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$. Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution can be approximated accurately by 
\begin{equation}
 f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean, white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)} +\xi^{(n)}(t) \ . 
	\label{eq:frequency_evolution}
\end{equation}
In Eq \ref{eq:frequency_evolution}, $f_{\rm em}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
\begin{equation}
	\langle \xi^{(n)}(t) \rangle = 0 \ ,
\end{equation}
\begin{equation}
	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle = [\sigma^{(n)}]^2 \delta(t - t') \ .
	\label{eq:xieqn}
\end{equation}
$[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Eq. \ref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments. \newline

@@

\section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e. 
\begin{equation}
	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
\end{equation}
with
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ ,
\end{equation}
and
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
\end{equation}
In Equation \eqref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
\begin{align}
	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ , 
\end{align}
where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer line of sight, i.e. $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$. We use this parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work. For a PTA dataset containing $N$ pulsars we then have $7 + 5N$ parameters to estimate. Typically the pulsar parameters are better constrained than the GW parameters; for example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information on the SMBHB GW source location.  \newline

In this section we present a new method to infer $\boldsymbol{\theta}$ and calculate the marginal likelihood (i.e. the model evidence). In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter. In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference of the model parameters and calculate the model evidence. Model selection and the specification of the null model is described in Section \ref{sec:model_selection}.

%The choice of prior distributions on the model parameters are described in

@@

\subsection{Nested Sampling}\label{sec:nested_sampling}
We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}}
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
In order to estimate the posterior distribution and the model evidence we use nested sampling \citep{Skilling} throughout this work. Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does this by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and then iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is the ability to compute the evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches. Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We  follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. The primary tunable parameter in nested sampling is $n_{\rm live}$, where a greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}(1/\sqrt{n_{\rm live}})$. However the computational runtime scales as $\mathcal{O}(n_{\rm live})$ and so one must make a trade-off between uncertainty and runtime. \cite{Ashton2022} offer a rule-of-thumb where the minimum number of live points should be greater than the number of static parameters. The results presented in this work are generally robust to the choice of $n_{\rm live}$, subject to the requirement of $n_{\rm live} > 7 + 5N$. Unless stated otherwise we take $n_{\rm live} = 500$ for all results presented in this work. \newline

@@

\subsection{Summary of workflow}
To summarise the above, the order of operations in our method for estimating the posteriors on the static parameters and calculating the marginal likelihood for the inference model is as follows:

\begin{enumerate}
	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\boldsymbol{\theta}_{\rm sample}$
	\item Pass $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
	\item Iterate over the data $\boldsymbol{Y}$ using the Kalman filter and obtain a single $\log \mathcal{L}$ value
	\item Return $\log \mathcal{L}$ to the nested sampler
	\item Nested sampler updates $\boldsymbol{\theta}$
	\item Repeat
\end{enumerate}
In Section \ref{sec:testing} we apply this workflow on a synthetic data set.

@@

For this work we generate synthetic data that is astrophysically representative as follows. We first specify a set of pulsars to make up our artificial PTA. We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA and the method naturally extends to other PTAs of the IPTA. For each pulsar we obtain the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$. The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Fig \ref{fig:pulsar_distrib}. The data is acquired via the Australia Telescope National Facility pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining unspecified pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$. We discussed in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$. Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ \citep{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars. For this work we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. In order to set a physically reasonable value for $[\sigma^{(n)}]^2$ we use two complementary approaches. The first approach uses the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} where the standard deviation in the pulsar TOA is given by:
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)&= \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T$ is the length of the observation span and for MSPs $\ln \alpha_1 \sim -20 $, $\alpha_2 \sim 1$, $\alpha_3 \sim 2$ and $\alpha_4 \sim 2.4$. Throughout this work we assume that all pulsars are observed with a weekly cadence i.e. $T = 1$ week. The frequency noise parameter can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as 
\begin{eqnarray}
	\sigma^{(n)} \sim \sigma_{\rm TOA}^{(n)} \frac{f_{\rm p}^{(n)}}{T} \ . \label{eq:sigmap_f}
\end{eqnarray}
For our synthetic NANOGrav PTA, the median $\sigma^{(n)}$ calculated in this way is $ = 4.28 \times 10^{-21} $ s. \newline

>>>>

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[Kimpson]{Kimpson$^{1}$, Melatos$^{2}$, O'Leary, Evans$^{3}$, Carlin others, etc. etc.%
\thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
%\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
\\
% List of institutions
$^{1}$School of Physics, University of Melbourne, Parkville, VIC 3010, Australia \\
$^{2}$OzGrav, University of Melbourne, Parkville, VIC 3010, Australia \\
$^{3}$Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia }

@@

% Abstract of the paper
\begin{abstract}	
Continuous nHz gravitational waves from individual supermassive black hole binaries may be detectable with Pulsar Timing Array observations. A novel search strategy is developed wherein intrinsic pulsar timing variations are tracked simultaneously with the modulation induced by a putative gravitational wave. A two-step procedure is applied within a state-space framework, where the model parameters are estimated with a Kalman filter, which provides a likelihood for nested sampling. This enables self-consistent inference of model parameters and calculation of the Bayesian evidence. It is shown via an astrophysically representative software injection campaign that the procedure can distinguish a gravitational wave from pure noise down to a strain of $h_0 \sim 4 \times 10^{15}$, given an inclination angle $\iota = 1.0$ rad. Full posterior distributions of model parameters are recovered. There is a bias in the estimates of some parameters, introduced by dropping the so-called `pulsar terms', which is analysed and discussed.
\end{abstract}

@@

\section{State-Space Formulation}\label{sec:model}
We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation. In this work we take the intrinsic state variable to be the pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local rest frame of the pulsar's centre of mass. A model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.  We take the measurement variable to be the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$.  The measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_measured}. The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array.
\subsection{Spin evolution} \label{sec:psr_frequency}
A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}. We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend. In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque. The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$. Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution can be approximated accurately by 
\begin{equation}
 f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean, white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)} +\xi^{(n)}(t) \ . 
	\label{eq:frequency_evolution}
\end{equation}
In Eq \ref{eq:frequency_evolution}, $f_{\rm em}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
\begin{equation}
	\langle \xi^{(n)}(t) \rangle = 0 \ ,
\end{equation}
\begin{equation}
	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle = [\sigma^{(n)}]^2 \delta(t - t') \ .
	\label{eq:xieqn}
\end{equation}
$[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Eq. \ref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments. \newline

@@

\section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar, and parameters of the GW source, i.e. 
\begin{equation}
	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
\end{equation}
with
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ , \label{eq:psrparams}
\end{equation}
and
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
\end{equation}
In Equation \eqref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
\begin{align}
	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ , 
\end{align}
where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer line of sight, i.e. $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$. We use this parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work. For a PTA dataset containing $N$ pulsars we then have $7 + 5N$ parameters to estimate. Typically the pulsar parameters are better constrained than the GW parameters; for example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information on the SMBHB GW source location.  \newline

In this section we present a new method to infer $\boldsymbol{\theta}$ and calculate the marginal likelihood (i.e. the model evidence). In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency, $f_{\rm m}^{(n)}(t)$, can be used to estimate the hidden state sequence, $f_{\rm p}^{(n)}(t)$, using a Kalman filter. In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to perform Bayesian inference of the model parameters and calculate the model evidence. Model selection and the specification of the null model is described in Section \ref{sec:model_selection}. A complete summary of the method is presented in Section \ref{sec:methodsummary}.

@@

\subsection{Nested Sampling}\label{sec:nested_sampling}
We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
In order to estimate the posterior distribution and the model evidence we use nested sampling \citep{Skilling} throughout this work. Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does this by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and then iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is the ability to compute the evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches. Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We  follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. The primary tunable parameter in nested sampling is $n_{\rm live}$, where a greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}(1/\sqrt{n_{\rm live}})$. However the computational runtime scales as $\mathcal{O}(n_{\rm live})$ and so one must make a trade-off between uncertainty and runtime. \cite{Ashton2022} offer a rule-of-thumb where the minimum number of live points should be greater than the number of static parameters. The results presented in this work are generally robust to the choice of $n_{\rm live}$, subject to the requirement of $n_{\rm live} > 7 + 5N$. Unless stated otherwise we take $n_{\rm live} = 1000$ for all results presented in this work. \newline

@@

\subsection{Summary of workflow}\label{sec:methodsummary}
To summarise the above, the order of operations in our method for estimating the posteriors on the static parameters and calculating the marginal likelihood for the inference model is as follows:

\begin{enumerate}
	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\boldsymbol{\theta}_{\rm sample}$
	\item Pass $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
	\item Iterate over the data $\boldsymbol{Y}$ using the Kalman filter and obtain a single $\log \mathcal{L}$ value
	\item Return $\log \mathcal{L}$ to the nested sampler
	\item Nested sampler updates $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$
	\item Repeat
\end{enumerate}
In Section \ref{sec:testing} we apply this workflow on a synthetic data set.

@@

For this work we generate synthetic data that is astrophysically representative as follows. We first specify a set of pulsars to make up our artificial PTA. We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA and the method naturally extends to other PTAs of the IPTA. For each pulsar we obtain the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$. The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Figure \ref{fig:pulsar_distrib}. The data is acquired via the Australia Telescope National Facility pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining unspecified pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$. We discussed in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$. Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ \citep{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars. For this work we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. In order to set a physically reasonable value for $[\sigma^{(n)}]^2$ we use two complementary approaches. The first approach uses the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} where the standard deviation in the pulsar TOA is given by:
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)&= \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T$ is the length of the observation span and for MSPs $\ln \alpha_1 \sim -20 $, $\alpha_2 \sim 1$, $\alpha_3 \sim 2$ and $\alpha_4 \sim 2.4$. Throughout this work we assume that all pulsars are observed with a weekly cadence i.e. $T = 1$ week. The frequency process noise parameter $\sigma^{(n)}$ can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as 
\begin{eqnarray}
	\sigma^{(n)} \sim \sigma_{\rm TOA}^{(n)} \frac{f_{\rm p}^{(n)}}{T} \ . \label{eq:sigmap_f}
\end{eqnarray}
For our synthetic NANOGrav PTA, the median $\sigma^{(n)}$ calculated in this way is $ = 4.28 \times 10^{-21} $ s. \newline

As a sanity check, we can also calculate $\sigma^{(n)}$ using a complementary numerical approach to the empirical best-fit model outlined above. We can directly solve the state equation Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a synthetic phase solution,
\begin{eqnarray}
	\xi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
\end{eqnarray}
and generate TOAs and phase residuals. We calibrate the noise amplitude $\sigma^{(n)}$  to generate phase residuals that qualitatively resemble empirical phase residuals measured from real pulsars. We obtain empirical phase residuals from the UTMOST pulsar timing program \citep{UTMOST} from the  Molonglo Observatory Synthesis Telescope \citep{Bailes2017PASA...34...45B}. For the purposes of validating the $\sigma^{(n)}$ inferred from Equation \eqref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real phase residuals visually. Whilst it would also be possible to e.g. define some quantitative loss function (e.g. mean-squared error) between the two solutions and solve the optimisation problem to infer $\sigma^{(n)}$, for our purposes this is excessive. It is satisfactory to simply qualitatively compare the two solutions as a useful sanity check. We emphasise that we are not overly concerned with calculating the maximally accurate values for $\sigma^{(n)}$, but instead some reasonable values for constructing representative synthetic data. This approach of generating synthetic phase solutions and visually comparing with empirical solutions is also followed in \cite{Vargas}. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for the NANOGrav pulsar J0030+0451. We can see that the synthetic and empirical residuals are qualitatively similar.
\begin{figure}
	\includegraphics[width=\columnwidth]{images/example_residuals_plot}
	\caption{Actual (left panels) and synthetic (right panels) phase residuals for NANOGrav millisecond pulsar J0030+0451. The actual residuals are obtained via UTMOST pulsar timing program \citep{UTMOST}. The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Equations \eqref{eq:sigmap}, \eqref{eq:sigmap_f}. For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals. }
	\label{fig:qualitative_compare}
\end{figure}
\subsubsection{Synthetic data generation}
We generate $N$ synthetic noisy timeseries - one for each pulsar - as follows:
